---
title: "Multilevel Models"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: slides_preamble.tex
 pdf_document: 
  keep_tex: true
  includes:
   in_header: slides_preamble.tex
bibliography: refs.bib
csl: apa.csl
---


```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, prompt = TRUE, warning = FALSE, message = FALSE, comment=NA)
library(lme4)
library(ggplot2)
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(nlme)
library(brms)

schools <- c('1308', '1946', '7276', '9198', '2917', '9104', '4042', '7635', '8188', '6464', '5838', '9158', '5815', '6808', '6578', '4350')
Df <- MathAchieve %>% 
  select(school = School,
         minority = Minority,
         sex = Sex,
         ses = SES,
         mathach = MathAch) %>% 
  filter(school %in% schools) %>% 
  mutate(school = as.character(school))

set.seed(10101)
mu <- rnorm(16, mean=500, sd=50)
f <- function(mu) rnorm(250, mean=mu, sd=25)
fake_rt <- do.call(cbind, lapply(mu, f)) %>% 
  as.data.frame() %>% 
  as_tibble() %>% 
  rename_all(funs(str_replace(., "V", "subject"))) %>% 
  gather(subject, reaction_time, subject1:subject16) 

#write_csv(fake_rt, '../data/fake_rt.csv')

```

# Multilevel data: Example 1

Reaction times[^1] for 16 subjects.

[^1]: This is fake data. Real reaction times would not look so normal. 

```{r}
ggplot(fake_rt,
       aes(x = reaction_time, fill = subject)
) + geom_histogram(col='white', binwidth = 10) +
  facet_wrap(~ subject) + guides(fill=F)
```

# Multilevel data: Example 2

Reaction time as a function of sleep deprivation.
```{r}
ggplot(sleepstudy,
       aes(x=Days, y=Reaction, col=Subject)
) + geom_point() +
  stat_smooth(method='lm', se=F) +
  facet_wrap(~Subject) +
  guides(col=F)
```

# Multilevel data: Example 3

Mathematical achievement as function of socio-economic status.
```{r}
ggplot(Df,
       aes(x=ses, y=mathach, col=school)
) + geom_point() +
  stat_smooth(method='lm', se=F) +
  facet_wrap(~school) +
  guides(col=F)
```




# Example: Multilevel model for reaction times

-   Consider we have reaction time data from $J$ subjects,
    $$\{x_{j1}, x_{j2}, x_{j3} \ldots x_{jn_j}\}_{j=1}^J.$$

-   A simple multilevel model for this data might be: $$\begin{aligned}
    x_{ji} &\sim N(\mu_j,\sigma^2),\quad\text{for $i \in \{1\ldots n_j\}$},\\
    \mu_j &\sim N(\theta,\tau^2), \quad\text{for $j \in \{1 \ldots J\}$}.\end{aligned}$$

-   In words, each $x_{ji}$ is drawn from a Gaussian with mean $\mu_j$
    and variance $\sigma^2$, and each $\mu_j$ is drawn from a Gaussian
    with mean $\theta$ and variance $\tau^2$.
    
- A Bayesian model will put a prior over $\theta$ and $\tau$,
  and infer the posterior over $\theta$, $\tau$, $\mu_1 \ldots \mu_J$,
  $\sigma^2$.

# Example: Multilevel model for reaction times

-   We can re-write $x_{ji} \sim N(\mu_j,\sigma^2)$ as
    $$x_{ji} = \mu_j + \epsilon_{ji},\quad \epsilon_{ji} \sim N(0,\sigma^2).$$

-   We can re-write $\mu_j\sim N(\theta,\tau^2)$ as
    $$\mu_j = \theta + \eta_j, \quad \eta_{j} \sim N(0,\tau^2).$$

-   The multilevel model can be re-written
    $$x_{ji} = \theta + \eta_j + \epsilon_{ji}\quad \epsilon_{ji} \sim N(0,\sigma^2), \eta_{j} \sim N(0,\tau^2).$$

-   This is often termed a *random-effects* model.

# Example: Multilevel model for reaction times


-   The variable $\theta$ denotes the global average reaction time.

-   The variables $\mu_1 \ldots \mu_j \ldots \mu_J$ are
    the subjects's average reaction times. 

-   The variables $\eta_1 \ldots \eta_j \ldots \eta_J$ are the
    offsets of each subject's average reaction time from the global
    average. Each $\eta_j = \mu_j - \theta$.

-   The variable $\sigma^2$ denotes the variance within any given
    subject.

-   The variable $\tau^2$ denotes the variance across subjects.

# Example: Multilevel model for reaction times

-   In the model just described, $\theta$ tells us the global average.

-   The variance $\tau^2$ tells us how much any given subject's average
    varies about $\theta$.

-   For example, 95% and 99% of the averages for individual subjects,
    will be in the ranges
    $$\theta \pm 1.96 \times \tau,\quad  \theta \pm 2.56 \times \tau,$$
    respectively.

-   Likewise, 95% and 99% of any given subject's reaction times, i.e.
    $x_{ji}$, will be in the ranges
    $$\theta + \eta_j   \pm 1.96 \times \sigma,\quad \theta + \eta_j  \pm 2.56 \times \sigma.$$



# Example: Multiple drivers, multiple cars

-   Let's say we want to measure the mpg of a given model of car (e.g. a
    Porsche 911).

-   Because any one car could vary from others of the same model, we
    have $K$ different examples of this model of car.

-   Likewise, because any one driver could affect the recorded mpg of
    the car he drives, we have $J$ different drivers.

-   We get each of the J drivers to drive each of the K cars, and record
    the mpg as $$y_{jk} = \text{mpg for driver $j$, car $k$}.$$

# Example: Multiple drivers, multiple cars

-   A multilevel model for this mpg experiment could be
    $$\begin{aligned}
    y_{jk} &\sim N(\mu_{j} + \nu_k,\sigma^2),\\
    \mu_{j} &\sim N(\phi,\tau^2)\\
    \nu_{k} &\sim N(\psi,\upsilon^2)\end{aligned}$$ which would work out
    as
    $$y_{jk} = \underbrace{\theta}_{\phi + \psi} + \eta_j + \zeta_k + \epsilon_{jk},$$
    with
    $$\eta_j \sim N(0,\tau^2),~ \zeta_k \sim N(0,\upsilon^2),~ \epsilon_{jk} \sim N(0,\sigma^2).$$

# Example: Multiple drivers, multiple cars

-   In this example, we have three sources of variation
    $$y_{jk} = \theta + \underbrace{\eta_j}_{\text{\tiny within driver}} + \underbrace{\zeta_k}_{\text{\tiny within car}} + \underbrace{\epsilon_{jk}}_{\text{\tiny within trial}},$$
    where $\tau^2$ gives the within driver variance, $\upsilon^2$ gives
    the within car variation, and $\sigma^2$ gives within trial
    variation.

-   The variable $\theta$ provides the average mpg for the car model
    (i.e. the Porsche 911)

-   The variables $\tau^2$, $\upsilon^2$ and
    $\sigma^2$ provide measures of the relative variation across in mpg
    drivers, cars and trials, respectively.

# Example: Reaction time and math achievement

-   In this problem, we have $J$ subject. For subject $j$, we have
    $n_j$ data points.

-   In observation $i$ from subject $j$, their number of days without sleep is $x_{ji}$ and the reaction time is $y_{ji}$.

-   A multilevel model for this data is $$\begin{aligned}
    y_{ji} &\sim N(\alpha_{j} + \beta_j x_{ji},\sigma^2),\\
    \alpha_j &\sim N(a,\tau_a^2),\\
    \beta_j &\sim N(b,\tau_b^2).\end{aligned}$$

# Example: Reaction time and math achievement

-   The model $$\begin{aligned}
    y_{ji} &\sim N(\alpha_{j} + \beta_j x_{ji},\sigma^2),\\
    \alpha_j &\sim N(a,\tau_a^2),\\
    \beta_j &\sim N(b,\tau_b^2),\end{aligned}$$ can be re-written
    $$y_{ji} = \underbrace{(a + \eta_j)}_{\text{ $\alpha_j$}} + \underbrace{(b + \zeta_j)}_{\text{ $\beta_j$}} x_{ji} + \epsilon_{ji},$$
    or
    $$y_{ji} = \underbrace{a + bx_{ji}}_{\text{Fixed effect}} + \underbrace{\eta_j + \zeta_j x_{ji}}_{\text{Random effect}} + \epsilon_{ji},$$
    where
    $$\eta_j \sim N(0,\tau_a^2),~ \zeta_j \sim N(0,\tau_b^2),~ \epsilon_j \sim N(0,\sigma^2).$$

    



# Example:  Reaction time and math achievement

-   In the model just described, $a$ and $b$ are the general regression
    coefficients.

-   The variance $\tau_a^2$ tells us how much variation in the intercept
    term there is across schools. The variance $\tau_b^2$ tells us how
    much variation in the slope term there is across schools.

-   For example, 95% and 99% of the intercepts for individual schools
    will be in the ranges
    $$a \pm 1.96 \times \tau_a,\quad  a \pm 2.56 \times \tau_a,$$
    respectively. Likewise, 95% and 99% of the slope terms for schools
    will be in the ranges
    $$b \pm 1.96 \times \tau_b,\quad b  \pm 2.56 \times \tau_b.$$

